# DBM analysis text output {#quick-start-dbm-text}



```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(RJafroc)
```

## TBA How much finished {#quick-start-dbm-text-how-much-finished}
50%


## Introduction {#quick-start-dbm-text-intro}
This chapter illustrates significance testing using the DBM method. 


## Analyzing the ROC dataset {#quick-start-dbm-text-analyze-dataset}

This illustrates the `StSignificanceTesting()` function. The significance testing method is specified as `"DBM"` and the figure of merit `FOM` is specified as "Wilcoxon". The embedded dataset `dataset03` is used.

```{r quick-start-dbm-text-significance-testing}
ret <- StSignificanceTesting(dataset03, FOM = "Wilcoxon", method = "DBM")
```             

## Explanation of the output {#quick-start-dbm-text-explanation}
The function returns a list with 5 members: 

* `FOMs`: figures of merit.
* `ANOVA`: ANOVA tables.
* `RRRC`: random-reader random-case analyses results.
* `FRRC`: fixed-reader random-case analyses results.
* `RRFC`" random-reader fixed-case analyses results.

Let us consider them individually. 

```{r}
str(ret$FOMs)
```             

* `FOMs` is a list of 3
    + `foms` is a [2x4] dataframe: the figure of merit for each of of the four observers in the two treatments.
    + `trtMeans` is a [2x1] dataframe: the average figure of merit over all readers for each treatment.
    + `trtMeanDiffs` a [1x1] dataframe: the difference(s) of the reader-averaged figures of merit for all different-treatment pairings. In this example, with only two treatments, there is only one different-treatment pairing.
    
```{r}
ret$FOMs$foms
ret$FOMs$trtMeans
ret$FOMs$trtMeanDiffs
```             


```{r}
str(ret$ANOVA)
```             

* ANOVA is a list of 4
    + `TRCanova` is a [8x3] dataframe: the treatment-reader-case ANOVA table, see below, where SS is the sum of squares, DF is the denominator degrees of freedom and MS is the mean squares, and T = treatment, R = reader, C = case, TR = treatment-reader, TC = treatment-case, RC = reader-case, TRC = treatment-reader-case.  
    + `VarCom` is a [6x1] dataframe: the variance components, see below, where `varR` is the reader variance, `varC` is the case variance, `varTR` is the treatment-reader variance, `varTC` is the treatment-case variance, `varRC` is the reader-case variance, and `varTRC` is the treatment-reader-case variance.
    + `IndividualTrt` is a [3x3] dataframe: the individual treatment variance components averaged over all readers, see below, where `msR` is the mean square reader, `msC` is the mean square case and `msRC` is the mean square reader-case.
    + `IndividualRdr` is a [3x5] dataframe: the individual reader variance components averaged over treatments, see below, where `msT` is the mean square treatment, `msC` is the mean square case and `msTC` is the mean square treatment-case.
    
```{r}
ret$ANOVA$TRCanova
ret$ANOVA$VarCom
ret$ANOVA$IndividualTrt
ret$ANOVA$IndividualRdr
```             

```{r}
str(ret$RRRC)
```             

* RRRC, a list of 3 containing results of random-reader random-case analyses
    + `FTtests`: is a [2x4] dataframe: results of the F-tests, see below, where `FStat` is the F-statistic and `p` is the p-value. The first row is the treatment effect and the second is the error term. 
    + `ciDiffTrt`: is a [1x7] dataframe: the confidence intervals between different-treatments, see below, where `StdErr` is the standard error of the estimate, `t` is the t-statistic and `PrGTt` is the p-value.
    + `ciAvgRdrEachTrt`: is a [2x5] dataframe: the confidence intervals for each treatment, averaged over all readers in the treatment, see below, where `CILower` is the lower 95% confidence interval and `CIUpper` is the upper 95% confidence interval.
    
```{r}
ret$RRRC$FTests
ret$RRRC$ciDiffTrt
ret$RRRC$ciAvgRdrEachTrt
```             


```{r}
str(ret$FRRC)
```             

* FRRC, a list of 4 containing results of fixed-reader random-case analyses
    + `FTtests`: is a [2x4] dataframe: results of the F-tests, see below.
    + `ciDiffTrt`: is a [1x7] dataframe: the confidence intervals between different-treatments, see below.
    + `ciAvgRdrEachTrt`: is a [2x5] dataframe: the confidence intervals for the average reader over each treatment
    + `ciDiffTrtEachRdr`: is a [4x7] dataframe: the confidence intervals for each different-treatment pairing for each reader. 
    
```{r}
ret$FRRC$FTests
ret$FRRC$ciDiffTrt
ret$FRRC$ciAvgRdrEachTrt
ret$FRRC$ciDiffTrtEachRdr
```             

```{r}
str(ret$RRFC)
```             

* RRFC, a list of 3 containing results of random-reader fixed-case analyses
    + `FTtests`: is a [2x4] dataframe: results of the F-tests, see below. 
    + `ciDiffTrt`: is a [1x7] dataframe: the confidence intervals between different-treatments, see below. 
    + `ciAvgRdrEachTrt`: is a [2x5] dataframe: the confidence intervals for the average reader over each  over each treatment.  

    
```{r}
ret$RRFC$FTests
ret$RRFC$ciDiffTrt
ret$RRFC$ciAvgRdrEachTrt
```             



## References {#quick-start-dbm-text-references}
